---
title: "Regression Project"
author: "JR Waggoner, Jose Baldomar, Andre Louie"
date: "April 10, 2019"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
library(dplyr)
library(ggplot2)
library(readr)
library(GGally)
library(broom)
library(jtools)
library(anytime)
library(car)
library(kableExtra)
library(EnvStats)
library(knitr)
library(corrplot)
```

``` {r importing the data,echo = FALSE}
# import the CSV
loan <- read.csv("LoanStats3a.csv")

#drop rows with NA for now, revisit missing data later
# loan <-loan[complete.cases(loan),]  
```

```{r Data manipulation,echo = FALSE}

#convert revolving util from string % to numeric decimal
loan$revol_util<- as.numeric(sub("%","",loan$revol_util))/100

#split term on "months", convert to numeric
loan$term<- as.numeric(sub("months","",loan$term))

#compute target dependent variable, total paid
loan <- loan %>%
  mutate(tot_paid = total_rec_prncp + total_rec_int)

# there are 15 purpose categories, -> factor 
loan$purpose <- factor(loan$purpose)

#convert character strings from form like '10-Dec' to form like '2010-dec'
# strings to be used with anytime function
loan$issue_d <- sub("11","2011",loan$issue_d,fixed = TRUE)
loan$issue_d <- sub("10","2010",loan$issue_d,fixed = TRUE)
loan$issue_d <- sub("9","2009",loan$issue_d,fixed = TRUE)
loan$issue_d <- sub("8","2008",loan$issue_d,fixed = TRUE)
loan$issue_d <- sub("7","2007",loan$issue_d,fixed = TRUE)

#convert issue_d from factor to to date-time 
loan$issue_d <-anytime(loan$issue_d)

# convert from '7-Dec' to '2007-Dec' for use with anytime()
loan$last_pymnt_d <- sub("16","2016",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("15","2015",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("14","2014",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("13","2013",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("12","2012",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("11","2011",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d  <- sub("10","2010",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d  <- sub("9","2009", loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d  <- sub("8","2008",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d  <- sub("7","2007",loan$last_pymnt_d ,fixed = TRUE)

#convert issue_d from factor to date-time 
loan$last_pymnt_d <-anytime(loan$last_pymnt_d)

```

```{r target variable,echo = FALSE}
#compute target dependent variable, total paid
loan <- loan %>%
  mutate(tot_paid = total_rec_prncp + total_rec_int)
```


### **Abstract**
Abstract text

### **Introduction**

  In researching this loan data set, our our intent is to better understand, and eventually model, the outcome of a loan.  More specifically, we wish to develop an understanding of those factors affect the overall performance or outcome of a LendingClub loan.  We expect a number of interesting business questions can be explored through our analysis, both boroad and narrow in scope.  We will explore loan outcomes as a component of estimating the total value of a loan portfolio.  More narrowly, we also wish to understand those factors that may have the most influence on the overall performance of a given loan.


### **Overview of the DataSet**
  The raw dataset is comprised of information describing all loans issued by LendingClub between 2007 and 2011. The data set can be found at  https://www.lendingclub.com/foliofn/rateDetail.action . From the raw data, a test set was created by first removing all columns that were entirely composed of missing values or contained the same value for each record.  Additionally, columns containing free text - title, description, etc., were removed along with redundant features that were a subset or component of another feature (sub-grade, funded amount,etc)  Finally, a target variable of 'total paid' was created by combining total principal and total interest, excluding any collection amounts or fees received.  Features relating to total amounts collected or recovered were also excluded from the test data set.

```{r exclude records, echo = FALSE}
#  pull three records from the dataframe, 
#   plus a fourth with a high loan amount

#copy row @ index 2000 to new DF
#test_case <- loan[c(13,1000,156,228),]
#delete row @ 2000 from original DF
#loan <- loan[-c(13,1000,156,228),]
```
  
### **Exploratory analysis**

##### **Variable Analysis**

```{r  paid dist, echo=FALSE}
# distribution of Total paid
ggplot(loan, aes(tot_paid))+
  geom_histogram(bins = 30)+
  ggtitle("Distribution of Total Paid")+
  xlab("Total Paid")
```

```{r loan amt dist,echo=FALSE}
#distribution of loan amounts
ggplot(loan,aes(loan_amnt))+
  geom_histogram(bins = 20)+
  ggtitle("Distribution of Loan Amounts")+
  xlab("Loan amount")
```

```{r int rate dist,echo=FALSE}
# distribution of interest rates
ggplot(loan,aes(int_rate))+
  geom_histogram(bins=20)+
  ggtitle("Distribution of Loan Interest Rates")+
  xlab("Interest Rate")
```

```{r income dist,echo=FALSE}
#distribution of of annual incomes
ggplot(loan,aes(annual_inc))+
  geom_histogram(bins = 20)+
  ggtitle(("Distribution of Borrower Annual Income"))+
  xlab("Annual Income")+
  xlim(0,200000)
```

```{r loan purpose, echo = FALSE}
# Count of loan purpose
ggplot(loan,aes(purpose))+
  geom_bar()+
  ggtitle("Count of Loans by purpose")+
  xlab("Purpose")+
  coord_flip()
```

```{r loans by date, echo = F}
ggplot(loan,aes(issue_d))+
  geom_histogram(bins = 48)+
  ggtitle("Count of Monthly Loan Issues")+
  xlab("Issue Date")+
  ylab("Count")
```

##### **Categorical Variable Analysis**

```{r repaid by purpose, echo = FALSE}
# boxplot of total repaid by purpose, with coordinates flipped
ggplot(loan, aes(purpose, tot_paid))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Purpose")+
  xlab("Purpose")+
  ylab("Total Paid")
```

```{r loan amount by purpose, echo = FALSE}
# boxplot of loan amount by purpose
ggplot(loan,aes(purpose,loan_amnt))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Loan Amount by Purpose")+
  xlab("Purpose")+
  ylab("Loan Amount")
```

```{r total paid by grade, echo = FALSE}
ggplot(loan,aes(grade,tot_paid))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by loan grade")+
  xlab("Loan Grade")+
  ylab("Total Paid")
```

```{r tota paid by home ownership, echo = FALSE}
ggplot(loan,aes(home_ownership,tot_paid))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Home Ownership")+
  xlab("Home Ownership")+
  ylab("Total Paid")
```

```{r total paid by verification status, echo = FALSE}
ggplot(loan,aes(verification_status,tot_paid))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Verification Status")+
  xlab("Verification Status")+
  ylab("Total Paid")
```


##### **Correlation Analysis**

**ggpairs PLOT IS DISABLED FOR EASIER KNITTING**
```{r pairwise plot, echo = FALSE}
numeric_loan <- loan %>%
  select(int_rate,installment,annual_inc,inq_last_6mths,
         mths_since_last_delinq,mths_since_last_record,open_acc,pub_rec,revol_bal,
         revol_util,recoveries,loan_amnt,tot_paid  )

#correlation plot
corrplot(cor(numeric_loan, use = "na.or.complete"),method = "ellipse")

#ggpairs for correlation numbers and distributions
#ggpairs(numeric_loan)

```

### **Model Identification and Evaluation** 
Our final OLS model was developed over 8 iterations, which we began by regressing on all available independent variables remaining after our intitial data cleanse.  Through five model iterations, we exlcuded the following variables that did not significantly contribute to our model (p > 0.7), and improved our model fit once removed (adjusted R^2): verification_status, zip_code, earliest_cr_line, total_acc, dti, last_credit_pull_d,grade, and delinq_2yrs.  loan_amnt was dropped due to its multicolinearity with the installment variable.  And, finally, we removed the state independent variable as its influence on total paid was minimal and the regression model could be made simpler by removing a 49-level factor variable. 
  Our 5th model iteration produced a set of residuals that failed both Non-constant variance and Shapiro-Wilk normailty tests:

```{r 6th model, echo = FALSE}
testmodel6 <- lm(tot_paid~term+int_rate+installment+grade+emp_length+
                   home_ownership+
                   annual_inc+issue_d+loan_status+purpose+
                   inq_last_6mths+mths_since_last_delinq+
                   mths_since_last_record+open_acc+pub_rec+revol_bal+revol_util+
                   last_pymnt_d+last_pymnt_amnt+recoveries
                 ,data=loan)
#summ(testmodel6) # adjusted R2 of 87.52
```

```{r ncv & Wilk teststest,echo = FALSE}
ncvTest(testmodel6)
shapiro.test(testmodel6$residuals)
```

A Box-Cox test of the output, total_paid, revealed a transformation between log and sqrt may help normalize residual variance:
```{r box cox,echo = FALSE}
box <- boxcox(testmodel6)
# boxcox test says lambda of somewhere between 0 and .5 may work for tot_paid
```

A cube-root transformation of our output produced residuals that passed the NCV test, but still failed the normailty test:

```{r model 7, echo = F}
testmodel7 <- lm((tot_paid^(1/3))~term+int_rate+installment+grade+emp_length+home_ownership+
                   annual_inc+issue_d+loan_status+purpose+
                   inq_last_6mths+mths_since_last_delinq+
                   mths_since_last_record+open_acc+pub_rec+revol_bal+revol_util+
                   last_pymnt_d+last_pymnt_amnt+recoveries
                 ,data=loan)

ncvTest(testmodel7) # p value of .726, non-constant variance is not an issue
```

A simiar process was followed for the independent variables where Box-Cox suggested a transformation would be appropriate.  Though normalizing a number of independent variables improved the fit of our model to an R^2 of 94.2, we were unable to produce a model that could pass both residiual tests.  

Summary of the final Model:
```{r transform input variables, echo = F}

#transform input variables based on estimated power output of boxcox tests
  # create new DF in which inputs will be transformed.
loantrans <- loan
loantrans$int_rate <- loantrans$int_rate^(.5678)
loantrans$installment <- loantrans$installment^(.2601)
#loantrans$annual_inc <- loantrans$annual_inc^(.076)
#loantrans$inq_last_6mths <- log(loantrans$inq_last_6mths)
loantrans$mths_since_last_delinq <- loantrans$mths_since_last_delinq^(.5837)
#loantrans$mths_since_last_record <- 1/(loantrans$mths_since_last_record^(.0688))
loantrans$open_acc <- loantrans$open_acc^(.286)
#loantrans$pub_rec <- 1/loantrans$pub_rec^(3)
loantrans$revol_bal <- loantrans$revol_bal^(.2773)
loantrans$revol_util <- loantrans$revol_util^(.3991)
loantrans$tot_paid <-loantrans$tot_paid^(.333333) # target outcome 

#copy rows to new DF
test_case <- loantrans[c(13,1000,156,228),]

#undo tot_paid transformation for predictions later on
test_case$tot_paid <- test_case$tot_paid^3

#delete rows from original DF
loantrans <- loantrans[-c(13,1000,156,228),]


```

```{r final model, echo = F}
testmodel10 <- lm(tot_paid~term+int_rate+installment+home_ownership+
                   annual_inc+issue_d+loan_status+purpose+
                   inq_last_6mths+open_acc+revol_bal+revol_util+
                   last_pymnt_d+last_pymnt_amnt+recoveries
                 ,data=loantrans)

summ(testmodel10) # adjusted R^2 of .942

modeldf1 <- augment(testmodel10)

#plot residiuals
ggplot(modeldf1,aes(.fitted,.resid))+
  geom_point()+
  #ylim(-15000,15000)+
  geom_line( y = 0, linetype = 2, color = "darkred")


ncvTest(testmodel10) #fail

#shapiro test with 5000 random samples, test needs 3 to 5000 samples

shapiro.test(sample(testmodel10$residuals,5000)) # fail
```


### **Model Application**
Applying the model to the four holdout cases produces the following estimates:
```{r predict for holdout case,echo = FALSE}
#apply model to test cases


predictions<- predict(testmodel10,test_case,interval = "predict")^3

combo <- test_case %>%
  select(tot_paid) 

pred_table <- merge(combo,predictions,by = 0, all = T)

pred_table <- pred_table%>%
  mutate("% Error" = (abs(tot_paid-fit)/tot_paid)*100)

colnames(pred_table) <- c("","Total Paid","Predicted","Lower PI","Upper PI","% Error")

```

```{r table,echo = F}

kable(pred_table) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

For these observations, the model does not appear to make  approporiate predictions for total amounts paid.  Total paid actual for the second observation, '13', falls outside of the 95% prediction interval.  Three of the observations had errors in excess of 20%, which is not appropriate given the potential business consequences of utilizing such erroneous estimates.   

### **Assumptions and Limitations**
Assumption/Limitation text

Limitations - 
greater economic factors and how they influence both the need for a loan, and what interest rates are charged.  
