---
title: "Regression Project"
author: "JR Waggoner, Jose Baldomar, Andre Louie"
date: "April 10, 2019"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(GGally)
```

``` {r importing the data,echo = FALSE}
# import the CSV
loandata <- read.csv("LoanStats3a.csv",stringsAsFactors = FALSE)
```

```{r subsetting the data,echo = FALSE}
# create a subset of data limited to those loans that failed + features 
loan <- loandata %>%
  filter(loan_status == "Charged Off") %>%
  mutate(pct_paid = total_rec_prncp/loan_amnt * 100) %>%
  select(loan_amnt,term,int_rate,inq_last_6mths,sub_grade,home_ownership,
         annual_inc,purpose,dti,pct_paid) 
```

```{r factor term, echo = FALSE}
#factor loan_term; 2 levels (3,5 years)
loan$term <- factor(loan$term, labels = c("3year","5year"))
```

```{r factor grades, echo = FALSE}
# factor loan grades info @ https://www.lendingclub.com/foliofn/rateDetail.action
loan$sub_grade <- factor(loan$sub_grade)
```

```{r factor ownership, echo = FALSE}
# factor home_ownership
loan$home_ownership <- factor(loan$home_ownership, levels = c("RENT","MORTGAGE","OWN"))
```

```{r factor purpose, echo = FALSE}
# there are 15 purpose categories, -> factor 
loan$purpose <- factor(loan$purpose)
```

### **Abstract**
Abstract text

### **Introduction**
Introduction text

### **Overview of the DataSet**
  The raw dataset is comprised of information describing all loans issued by LendingClub between 2007 and 2011. The data set can be found at 
  https://www.lendingclub.com/foliofn/rateDetail.action . A more focused data set was then created by selecting all records of failed loans,(i.e. where loan_status was recorded as "Charged Off",) and the following features:

  - **loan_amnt** - The original principal loan amount.
  + **term - Term** length of the loan: 36 or 60 months.
  + **int_rate** - Interest rate of the loan.
  + **inq_last_6mths** - # of inquiries on the borrowers credit report in the last 6 months
  + **sub_grade** - The loan subgrade assigned by Lending club: A1 (best) to G5 (worst.)
  + **home_ownership** - The borrowers ownership status: Rent, Mortgag or Own.
  + **annual_inc** - The borrowers annual income in USD.
  + **purpose** - General purpose for the loan: 1 of 15 categories.
  + **dti** - The borrower's debt-to-income ratio
  + **pct_paid** - percent of loan principal paid: total principal paid/loan amount.
  
### **Exploratory analysis**
Exploratory text

##### **Distributions**

```{r  paid dist, echo=FALSE}
# distribution of  % paid
ggplot(loan, aes(pct_paid))+
  geom_histogram(bins = 30)+
  ggtitle("Distribution of % Paid")+
  xlab("% Paid")
```

```{r loan amt dist,echo=FALSE}
#distribution of loan amounts
ggplot(loan,aes(loan_amnt))+
  geom_histogram(bins = 7)+
  ggtitle("Distribution of Loan Amounts")+
  xlab("Loan amount")
```

```{r int rate dist,echo=FALSE}
# distribution of interest rates
ggplot(loan,aes(int_rate))+
  geom_histogram(bins=20)+
  ggtitle("Distribution of Loan Interest Rates")+
  xlab("Interest Rate")
```


```{rdti dist,echo=FALSE}
# DTI distribution
ggplot(loan, aes(dti))+
  geom_histogram()+
  ggtitle("Distribution of Borrower DTI")+
  xlab("DTI")
```

```{r income dist,echo=FALSE}
#distribution of of annual incomes
ggplot(loan,aes(annual_inc))+
  geom_histogram(bins = 50)+
  ggtitle(("Distribution of Borrower Annual Income"))+
  xlab("Annual Income")
```

```{r loan purpose, echo = FALSE}
# Count of loan purpose
ggplot(loan,aes(purpose))+
  geom_bar()+
  ggtitle("Count of Loans by purpose")+
  xlab("Purpose")+
  coord_flip()
```

##### **Categorical Variables**

```{r repaid by purpose, echo = FALSE}
# boxplot of % repaid by purpose, with coordinates flipped
ggplot(loan, aes(purpose, pct_paid))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("% Paid by Purpose")+
  xlab("% Paid")+
  ylab("Purpose")
```

```{r repaid by grade, echo = FALSE}
# boxplot plot of loan grade vs.pct_paid
ggplot(loan,aes(sub_grade,pct_paid))+
  geom_boxplot()+
  ggtitle("% Paid by Sub-grade")+
  xlab("Loan Sub-Grade")+
  ylab("% Paid")
```

```{r repaid by term, echo = FALSE}
# boxplot of loanterm  vs. pct paid
ggplot(loan,aes(term,pct_paid)) +
  geom_boxplot()+
  ggtitle("% Paid by Loan Term")+
  xlab("Loan Term")+
  ylab("% Paid")
```

```{r repaid by ownership, echo = FALSE}
# boxplot of home ownership to pct_paid
ggplot(loan,aes(home_ownership,pct_paid))+
  geom_boxplot()+
  ggtitle("% Paid by Home Ownership")+
  xlab("Home Ownership")+
  ylab("% Paid")
```

```{r loan amount by purpose, echo = FALSE}
# boxplot of loan amount by purpose
ggplot(loan,aes(purpose,loan_amnt))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Loan Amount by Purpose")+
  xlab("Purpose")+
  ylab("Loan Amount")
```

##### **Correlation Analysis**
```{r pairwise plot, echo = FALSE}
#subset of only numeric vaiables
numeric_loan <- loan %>%
  select(loan_amnt,int_rate,inq_last_6mths,annual_inc,dti,pct_paid)

#pairwise ploits of numeric values  
ggpairs(numeric_loan)
```

### **Model Identification and Evaluation** 
Model Identification text


### **Model Application**
Model application text

### **Assumptions and Limitations**
Assumption/Limitation text


