---
title: "LendingClub Loan Data Analysis"
author: "JR Waggoner, Jose Baldomar, Andre Louie"
date: "April 26, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
library(dplyr)
library(ggplot2)
library(readr)
library(GGally)
library(broom)
library(jtools)
library(anytime)
library(car)
library(kableExtra)
library(EnvStats)
library(knitr)
library(corrplot)
library(caret)
library(tsensembler)
```

``` {r importing the data,echo = FALSE}
# import the CSV
loan <- read.csv("LoanStats3a.csv")

#drop rows with NA for now, revisit missing data later
# loan <-loan[complete.cases(loan),]  
```

```{r Data manipulation,echo = FALSE}

#convert revolving util from string % to numeric decimal
loan$revol_util<- as.numeric(sub("%","",loan$revol_util))/100

#split term on "months", convert to numeric
loan$term<- as.numeric(sub("months","",loan$term))

#compute target dependent variable, total paid
loan <- loan %>%
  mutate(tot_paid = total_rec_prncp + total_rec_int)

# there are 15 purpose categories, -> factor 
loan$purpose <- factor(loan$purpose)

#convert character strings from form like '10-Dec' to form like '2010-dec'
# strings to be used with anytime function
loan$issue_d <- sub("11","2011",loan$issue_d,fixed = TRUE)
loan$issue_d <- sub("10","2010",loan$issue_d,fixed = TRUE)
loan$issue_d <- sub("9","2009",loan$issue_d,fixed = TRUE)
loan$issue_d <- sub("8","2008",loan$issue_d,fixed = TRUE)
loan$issue_d <- sub("7","2007",loan$issue_d,fixed = TRUE)

#convert issue_d from factor to to date-time 
loan$issue_d <-anytime(loan$issue_d)

# convert from '7-Dec' to '2007-Dec' for use with anytime()
loan$last_pymnt_d <- sub("16","2016",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("15","2015",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("14","2014",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("13","2013",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("12","2012",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d <- sub("11","2011",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d  <- sub("10","2010",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d  <- sub("9","2009", loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d  <- sub("8","2008",loan$last_pymnt_d ,fixed = TRUE)
loan$last_pymnt_d  <- sub("7","2007",loan$last_pymnt_d ,fixed = TRUE)

#convert issue_d from factor to date-time 
loan$last_pymnt_d <-anytime(loan$last_pymnt_d)

```

```{r target variable,echo = FALSE}
#compute target dependent variable, total paid
loan <- loan %>%
  mutate(tot_paid = total_rec_prncp + total_rec_int)
```

### **Abstract**
  Between 2007 and 2011, LendingClub issued over 42,000 personal loans to borrowers.  To better understand loan performance in terms of total paid by a borrower, we conducted a multiple regression analysis of the loan data.  Our results illustrate the difficulty in modeling loan performance with multiple regression. In conclusion, we learned that LendingClub's loan data was too inconsistent to be used in a business context.  For future analysis, we may consider modeling with methods more appropriate for the dataset and adding additional sources of information.

### **Introduction**
  In researching this loan data set, our intent is to better understand, and eventually model, the outcome of a loan.  More specifically, we wish to develop an understanding of the factors that affect the overall performance or outcome of a LendingClub loan.  We expect a number of interesting business questions can be explored through our analysis, both boroad and narrow in scope.  We will explore loan outcomes as a component of estimating the total value of a loan portfolio.  More narrowly, we also wish to understand those factors that may have the most influence on the overall performance of a given loan.
  
### **Overview of the DataSet**
  The raw dataset is comprised of information describing all loans issued by LendingClub between 2007 and 2011. The data set can be found at  https://www.lendingclub.com/foliofn/rateDetail.action . From the raw data, a test set was created by first removing all columns that were entirely composed of missing values or contained the same value for each record.  Additionally, columns containing free text - title, description, etc., were removed along with redundant features that were a subset or component of another feature (sub-grade, funded amount,etc)  Finally, a target variable of 'total paid' was created by combining total principal and total interest, excluding any collection amounts or fees received.  Features relating to total amounts collected or recovered were also excluded from the test data set.

##### **Description of Variables**
  * annual_inc (numeric): The self-reported annual income provided by the borrower during registration  
  * emp_length (factor): Employment length in years
  * grade (factor): Lending Club's assigned loan grade
  * home_ownership (factor): The home ownership status provided by the borrower during registration
  * inq_last_6mths (numeric): The number of inquiries in past 6 months excluding auto and mortgage inquiries  
  * installment (numeric): The monthly payment owed by the borrower if the loan originates  
  * int_rate (numeric): Interest Rate on the loan
  * issue_d (date): The month which the loan was funded
  * last_pymnt_amnt (numeric): Last total payment amount received
  * last_pymnt_d (date): Last month payment was received  
  * loan_amnt (numeric): The listed amount of the loan applied for by the borrower. If at some point in time, the credit           department reduces the loan amount, then it will be reflected in this value  
  * loan_status (factor): Current status of the loan  
  * mths_since_last_delinq (numeric): The number of months since the borrower's last delinquency
  * mths_since_last_record (numeric): The number of months since the last public record
  * open_acc (numeric): The number of open credit lines in the borrower's credit file
  * pub_rec (numeric): The number of derogatory public records
  * purpose (factor): A category provided by the borrower for the loan request
  * recoveries (numeric): post charge off gross recovery
  * revol_bal (numeric): Total credit revolving balance
  * revol_util (numeric): Revolving line utilization rate, or the amount of credit the borrower is using relative to all           available revolving credit
  * term (factor): The number of payments on the loan
  * tot_paid (numeric): The sum of total principal and total interest, excluding any collection amounts or fees received
  * verification_status (factor): Indicates if the borrowers' income was verified by LC, not verified, or if the income           source was verified

```{r exclude records, echo = FALSE}
#  pull three records from the dataframe, 
#   plus a fourth with a high loan amount

#copy row @ index 2000 to new DF
#test_case <- loan[c(13,1000,156,228),]
#delete row @ 2000 from original DF
#loan <- loan[-c(13,1000,156,228),]
```
  
### **Exploratory analysis**

##### **Variable Analysis**
  Most of our variables have a right or positively skewed distribution. Respective to annual income and total loans paid, it makes sense that these observations have a greater mean than median. For example, from annnual income standpoint most borrowers or households typically make less than 100,000 with some borrowers quite a bit more ranging from 100,000 to 200,000.  
```{r  income dist, echo=FALSE}
# distribution of annual incomes
ggplot(loan,aes(annual_inc))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_histogram(bins = 20, color="black", fill="gray65")+
  ggtitle(("Borrower Annual Income"))+
  xlab("Annual Income")+
  xlim(0,200000)+
  ylab("Loan Count")
```

```{r installment dist,echo=FALSE}
# distribution of installment
ggplot(loan,aes(installment))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_histogram(bins=20, color="black", fill="gray65")+
  ggtitle("Installment")+
  xlab("Installment")+
  ylab("Loan Count")
```

```{r int rate dist,echo=FALSE}
# distribution of interest rates
ggplot(loan,aes(int_rate))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_histogram(bins=20, color="black", fill="gray65")+
  ggtitle("Loan Interest Rates")+
  xlab("Interest Rate")+
  ylab("Loan Count")
```

```{r loan amt dist,echo=FALSE}
# distribution of loan amounts 
ggplot(loan,aes(loan_amnt))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_histogram(bins = 20, color="black", fill="gray65")+
  ggtitle("Loan Amounts")+
  xlab("Loan amount")+
  ylab("Loan Count")
```

```{r months since last delinquency dist, echo = F}
# distribution of months since last delinquency
ggplot(loan, aes(mths_since_last_delinq))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_histogram(bins=30, color="black", fill="gray65")+
  ggtitle("Months Since Last Delinquency")+
  xlab("Months")+
  ylab("Loan Count")
```

```{r open accounts dist, echo = FALSE}
# distribution of open accounts
ggplot(loan,aes(open_acc))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_bar(color="black", fill="gray65")+
  ggtitle("Open Credit Line Accounts")+
  xlab("Open Accounts")+
  ylab("Loan Count")
```

```{r paid dist, echo = FALSE}
# distribution of total paid
ggplot(loan, aes(tot_paid))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_histogram(bins = 30, color="black", fill="gray65")+
  ggtitle("Total Paid")+
  xlab("Total Paid")+
  ylab("Loan Count")
```

##### **Categorical Variable Analysis**
  Looking at our boxplots in the analysis, it is clear to see that outliers are present in each one of our visuals. This reflects the point above about how many of our distributions are positively skewed. As we will further investgate, it is possible that data with this many outliers can cause problems from a constant variance and normailty perspective. 

```{r paid by purpose, echo = FALSE}
# boxplot of total paid by purpose, with coordinates flipped 
ggplot(loan, aes(purpose, tot_paid))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Purpose")+
  xlab("Purpose")+
  ylab("Total Paid")
```

```{r loan amount by purpose, echo = FALSE}
# boxplot of loan amount by purpose 
ggplot(loan,aes(purpose,loan_amnt))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Loan Amount by Purpose")+
  xlab("Purpose")+
  ylab("Loan Amount")
```

```{r total paid by grade, echo = FALSE}
# boxplot of total paid by grade
ggplot(loan,aes(grade,tot_paid))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Loan Grade")+
  xlab("Loan Grade")+
  ylab("Total Paid")
```

```{r tota paid by home ownership, echo = FALSE}
# boxplot of total paid by home ownership
ggplot(loan,aes(home_ownership,tot_paid))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Home Ownership")+
  xlab("Home Ownership")+
  ylab("Total Paid")

```

```{r total paid by employment length, echo = FALSE}
# boxplot of total paid by employment length
ggplot(loan,aes(emp_length,tot_paid))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Employment Length")+
  xlab("Employment Length")+
  ylab("Total Paid")
```

```{r total paid by verification status, echo = FALSE}
# boxplot of total paid by verification status
ggplot(loan,aes(verification_status,tot_paid))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Verification Status")+
  xlab("Verification Status")+
  ylab("Total Paid")
```

```{r total paid by loan status, echo = FALSE}
# boxplot of total paid by loan status
ggplot(loan,aes(loan_status,tot_paid))+
  ggtitle("Use theme(plot.title = element_text(hjust = 0.5)) to center") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("Total Paid by Loan Status")+
  xlab("Loan Status")+
  ylab("Total Paid")
```

##### **Correlation Analysis**
  In our correlation anlaysis, there is potential multicollinearity that can be spotted between two independent variables, installment and loan amount. If multicollinearity is indeed detected as we conduct our models, then we will need to remove one of these variables. 

**ggpairs PLOT IS DISABLED FOR EASIER KNITTING**
```{r pairwise plot, echo = FALSE}
numeric_loan <- loan %>%
  dplyr::select(int_rate,installment,annual_inc,inq_last_6mths,open_acc,
                pub_rec,revol_bal,revol_util,loan_amnt,tot_paid)

#correlation plot
corrplot(cor(numeric_loan, use = "na.or.complete"),method = "ellipse")

#ggpairs for correlation numbers and distributions
#ggpairs(numeric_loan)

```

### **Model Identification and Evaluation** 
  Our final OLS model was developed over 8 iterations, which we began by regressing on all available independent variables remaining after our intitial data cleanse.  Through five model iterations, we exlcuded the following variables that did not significantly contribute to our model (p > 0.7), and improved our model fit once removed (adjusted R^2): verification_status, zip_code, earliest_cr_line, total_acc, dti, last_credit_pull_d,grade, and delinq_2yrs. As proven true from our correlation analysis, the independent variable loan_amnt was dropped due to its multicolinearity with the installment. Lastly, we removed the state independent variable as its influence on the fit of the model was minimal, and the regression model could be made more manageable by removing a 49-level factor variable. 
  Our 6th model iteration produced a set of residuals that failed both Non-constant variance and Shapiro-Wilk normailty tests:

```{r 6th model, echo = FALSE}
testmodel6 <- lm(tot_paid~term+int_rate+installment+grade+emp_length+
                   home_ownership+
                   annual_inc+issue_d+loan_status+purpose+
                   inq_last_6mths+mths_since_last_delinq+
                   mths_since_last_record+open_acc+pub_rec+revol_bal+revol_util+
                   last_pymnt_d+last_pymnt_amnt+recoveries
                 ,data=loan)
#summ(testmodel6) # adjusted R2 of 87.52
```

```{r ncv & Wilk teststest,echo = FALSE}
ncvTest(testmodel6)
shapiro.test(testmodel6$residuals)
```

A Box-Cox test of the output, total_paid, revealed a transformation approximately between log(Y) and sqrt(Y) may help normalize residual variance:
```{r box cox,echo = FALSE}
box <- boxcox(testmodel6)
# boxcox test says lambda of somewhere between 0 and .5 may work for tot_paid
```

A cube-root transformation of our output produced residuals that passed the NCV test, but still failed the test of normality:

```{r model 7, echo = F}
testmodel7 <- lm((tot_paid^(1/3))~term+int_rate+installment+grade+emp_length+home_ownership+
                   annual_inc+issue_d+loan_status+purpose+
                   inq_last_6mths+mths_since_last_delinq+
                   mths_since_last_record+open_acc+pub_rec+revol_bal+revol_util+
                   last_pymnt_d+last_pymnt_amnt+recoveries
                 ,data=loan)

ncvTest(testmodel7) # p value of .726, non-constant variance is not an issue
```

A simiar process was followed for the independent variables where Box-Cox suggested a transformation would be appropriate.  Though normalizing a number of independent variables improved the fit of our model to an R^2 of 94.2, we were unable to produce a model that could pass both residiual tests.  

Summary of the final Model:
```{r transform input variables, echo = F}

#transform input variables based on estimated power output of boxcox tests
  # create new DF in which inputs will be transformed.
loantrans <- loan
loantrans$int_rate <- loantrans$int_rate^(.5678)
loantrans$installment <- loantrans$installment^(.2601)
#loantrans$annual_inc <- loantrans$annual_inc^(.076)
#loantrans$inq_last_6mths <- log(loantrans$inq_last_6mths)
loantrans$mths_since_last_delinq <- loantrans$mths_since_last_delinq^(.5837)
#loantrans$mths_since_last_record <- 1/(loantrans$mths_since_last_record^(.0688))
loantrans$open_acc <- loantrans$open_acc^(.286)
#loantrans$pub_rec <- 1/loantrans$pub_rec^(3)
loantrans$revol_bal <- loantrans$revol_bal^(.2773)
loantrans$revol_util <- loantrans$revol_util^(.3991)
loantrans$tot_paid <-loantrans$tot_paid^(.333333) # target outcome 

#copy test rows to new DF
test_case <- loantrans[c(13,1000,156,228),]

#undo tot_paid transformation for predictions 
test_case$tot_paid <- test_case$tot_paid^3

#delete test rows from original DF
loantrans <- loantrans[-c(13,1000,156,228),]


```

```{r final model, echo = F}
testmodel10 <- lm(tot_paid~term+int_rate+installment+home_ownership+
                   annual_inc+issue_d+loan_status+purpose+
                   inq_last_6mths+open_acc+revol_bal+revol_util+
                   last_pymnt_d+last_pymnt_amnt+recoveries
                 ,data=loantrans)

summ(testmodel10) # adjusted R^2 of .942

modeldf1 <- augment(testmodel10)

#plot residiuals
ggplot(modeldf1,aes(.fitted,.resid))+
  geom_point()+
  #ylim(-15000,15000)+
  geom_line( y = 0, linetype = 2, color = "darkred")


ncvTest(testmodel10) #fail

#shapiro test with 5000 random samples, test needs 3 to 5000 samples

shapiro.test(sample(testmodel10$residuals,5000)) # fail
```


### **Model Application**
Applying the model to the four holdout cases produces the following estimates:
```{r predict for holdout case,echo = FALSE}
#apply model to test cases


predictions<- predict(testmodel10,test_case,interval = "predict")^3

combo <- test_case %>%
  select(tot_paid) 

pred_table <- merge(combo,predictions,by = 0, all = T)

pred_table <- pred_table%>%
  mutate("% Error" = (abs(tot_paid-fit)/tot_paid)*100)

colnames(pred_table) <- c("","Total Paid","Predicted","Lower PI","Upper PI","% Error")

```

```{r table,echo = F}

kable(pred_table) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

For these observations, the model does not appear to make  approporiate predictions for total amounts paid.  Total paid actual for the second observation, '13', falls outside of the 95% prediction interval.  Three of the observations had errors in excess of 20%.

Applying the model to the training data and analyzing prediction errors produces an RMSE of :

```{r RMSE, echo = F}
mape_pred <- predict(testmodel10)^3

rmse(testmodel10$model$tot_paid^3,mape_pred)
```

Given the potential business consequences of utilizing such erroneous estimates, it would not be approporiate to apply this model to new or similar loan data.    

### **Assumptions and Limitations**
  Because the data was gathered and prepared by LendingClub, we were limited in our ability understand certain data points and, more importantly, handle missing information in the data set.  There are a number of borrower credit-related variables that were entirely blank that we feel could have helped build a better model of the total amount paid on a loan.  We also feel that it could be beneficial to bring in outside economic data, as we have hypothesized that economic factors may contribute to why a loan is taken out, its attributes, and how it eventually performs.  In the future, it would also be worth considering how LendingClub's growing popularity as a lending platform affects not only their lending frequency, but also the nature and performance of the loans themselves.    

  The first, and most broad, assumption we made was that the eventual outcome of a loan in terms of total paid could be described by a simple, multiple linear regression model, which does not appear to be the case.  The data may be more suitable for  tree-type or classification models, or for a binary outcome such as whether or not a loan was fully paid off.  

