---
title: "Regression Project"
author: "JR Waggoner, Jose Baldomar, Andre Louie"
date: "April 10, 2019"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(GGally)
```

``` {r importing the data,echo = FALSE}
# import the CSV
loandata <- read.csv("LoanStats3a.csv",stringsAsFactors = FALSE)
```

```{r subsetting the data,echo = FALSE}
# create a subset of data limited to those loans that failed + features 
loan <- loandata %>%
  filter(loan_status == "Charged Off") %>%
  mutate(pct_paid = total_rec_prncp/loan_amnt * 100) %>%
  select(loan_amnt,term,int_rate,inq_last_6mths,sub_grade,home_ownership,
         annual_inc,purpose,dti,pct_paid) 
```

```{r factor term, echo = FALSE}
#factor loan_term; 2 levels (3,5 years)
loan$term <- factor(loan$term, labels = c("3year","5year"))
```

```{r factor grades, echo = FALSE}
# factor loan grades info @ https://www.lendingclub.com/foliofn/rateDetail.action
loan$sub_grade <- factor(loan$sub_grade)
```

```{r factor ownership, echo = FALSE}
# factor home_ownership
loan$home_ownership <- factor(loan$home_ownership, levels = c("RENT","MORTGAGE","OWN"))
```

```{r factor purpose, echo = FALSE}
# there are 15 purpose categories, -> factor 
loan$purpose <- factor(loan$purpose)
```

### Abstract
Abstract text

### Introduction
Introduction text

### Overview of the DataSet
  The raw dataset is comprised of information describing all loans issued by LendingClub between 2007 and 2011. The data set can be found at 
  https://www.lendingclub.com/foliofn/rateDetail.action . A more focused data set was then created by selecting all records of failed loans,(i.e. where loan_status was recorded as "Charged Off",) and the following features:

  - **loan_amnt** - The original principal loan amount.
  + **term - Term** length of the loan: 36 or 60 months.
  + **int_rate** - Interest rate of the loan.
  + **inq_last_6mths** - # of inquiries on the borrowers credit report in the last 6 months
  + **sub_grade** - The loan subgrade assigned by Lending club: A1 (best) to G5 (worst.)
  + **home_ownership** - The borrowers ownership status: Rent, Mortgag or Own.
  + **annual_inc** - The borrowers annual income in USD.
  + **purpose** - General purpose for the loan: 1 of 15 categories.
  + **dti** - The borrower's debt-to-income ratio
  + **total_pymnt** - total principal + interest received before the loan was charged off.
  + **total_rec_prncp** - total principal received before the loan was charged off.
  + **pct_paid** - percent of loan principal paid: total principal paid/loan amount.
  
### Exploratory analysis
Exploratory text


##### Categorical Variables

```{r repaid by purpose, echo = FALSE}
# boxplot of % repaid by purpose, with coordinates flipped
ggplot(loan, aes(purpose, pct_paid))+
  geom_boxplot()+
  coord_flip()+
  ggtitle("% Paid by Purpose")+
  xlab("% Paid")+
  ylab("Purpose")
```

```{r repaid by grade, echo = FALSE}
# boxplot plot of loan grade vs.pct_paid
ggplot(loan,aes(sub_grade,pct_paid))+
  geom_boxplot()+
  ggtitle("% Paid by Sub-grade")+
  xlab("Loan Sub-Grade")+
  ylab("% Paid")
```

```{r repaid by term, echo = FALSE}
# boxplot of loanterm  vs. pct paid
ggplot(loan,aes(term,pct_paid)) +
  geom_boxplot()+
  ggtitle("% Paid by Loan Term")+
  xlab("Loan Term")+
  ylab("% Paid")
```

```{r repaid by ownership, echo = FALSE}
# boxplot of home ownership to pct_paid
ggplot(loan,aes(home_ownership,pct_paid))+
  geom_boxplot()+
  ggtitle("% Paid by Home Ownership")+
  xlab("Home Ownership")+
  ylab("% Paid")
```

##### Pair Plots
```{r pairwise plot, echo = FALSE}
#subset of only numeric vaiables
numeric_loan <- loan %>%
  select(loan_amnt,int_rate,inq_last_6mths,annual_inc,dti,pct_paid)

#pairwise ploits of numeric values  
ggpairs(numeric_loan)
```

### Model Identification and Evaluation 
Model Identification text


### Model Application
Model application text

### Assumptions and Limitations
Assumption/Limitation text


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
